---
title: "STAT406 - Lecture 7 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lecture slides

The lecture slides are [here](STAT406-17-lecture-7-preliminary.pdf).

## Comparing LASSO with Ridge Regression on the air pollution data

Let us compare the Ridge Regression and LASSO fits to the
air pollution data. Of course, by *the Ridge Regression fit*
and *the LASSO fit* we mean the fit obtained with the
optimal value of the penalty constant chosen in terms
of the corresponding estimated MSPE (which is in 
general estimated using K-fold cross validation). 

We first load the data and use `cv.glmnet()` with 
`alpha = 0` to select an **approximately optimal** 
Ridge Regression fit (what makes the calculation 
below **only approximately** optimal?).

```{r comparing.airp, fig.width=5, fig.height=5, warning=FALSE, message=FALSE}
airp <- read.table('../Lecture1/rutgers-lib-30861_CSV-1.csv', header=TRUE, sep=',')
y <- as.vector(airp$MORT)
xm <- as.matrix(airp[, names(airp) != 'MORT'])
library(glmnet)
lambdas <- exp( seq(-3, 10, length=50))
# Ridge Regression
set.seed(23)
air.l2 <- cv.glmnet(x=xm, y=y, lambda=lambdas, nfolds=5, alpha=0, 
                 family='gaussian', intercept=TRUE)
plot(air.l2)
```

The plot above is included for illustration purposes only. 
Similarly, we now compute an approximately optimal LASSO fit, 
and look at the curve of estimated MSPEs:

```{r airp.lasso, fig.width=5, fig.height=5, warning=FALSE, message=FALSE}
# LASSO
set.seed(23)
air.l1 <- cv.glmnet(x=xm, y=y, lambda=lambdas, nfolds=5, alpha=1, 
                 family='gaussian', intercept=TRUE)
plot(air.l1)
```

It is interesting to compare the corresponding estimated regression coefficients, 
so we put them side by side in two columns:
```{r airp.corr.groups} 
cbind(round(coef(air.l2, s='lambda.min'), 3),
round(coef(air.l1, s='lambda.min'), 3))
```

Note how several of them are relatively similar, but LASSO includes fewer of them.
A possible explanation for this is the particular correlation structure among the
explanatory variables. More specifically, when groups of 
correlated covariates are present, 
LASSO tends to choose only one of them, whereas Ridge Regression will tend 
to keep all of them. For a formal statement see Lemma 2 of

> Zou, H. and Hastie, T. (2005). Regularization and variable selection via the elastic net. 
> *Journal of the Royal Statistical Society: Series B (Statistical Methodology)*, 
> **67**(2), 301-320. 
> [DOI: 10.1111/j.1467-9868.2005.00503.x](http://dx.doi.org/10.1111/j.1467-9868.2005.00503.x)

It is important to note here that the above observations regarding the Ridge Regression
and LASSO fits trained on the air pollution data should be made on a more 
reliable (more stable, less variable) choice of penalty parameter. For example,
we may want to run the above 5-fold CV experiments several times and take the
average of the estimated optimal penalty parameters. To simplify the presentation
we do not purse this here, but it may be a very good exercise for the reader to
do so.

The following heatmap of the pairwise correlations among explanatory variables
reveals certain patterns that may be used to explain the difference
mentioned above. Note that in this visualization method variables were
grouped ("clustered") according to their pairwise correlations in order to
improve the interpretability of the plot. We will see later in this course
the particular clustering method used here (hierarchical clustering).

```{r airp.correlations, fig.width=5, fig.height=5, warning=FALSE, message=FALSE}
library(ggcorrplot)
ggcorrplot(cor(xm), hc.order = TRUE, outline.col = "white")
```

<!-- # # https://briatte.github.io/ggcorr/ -->
<!-- # source('https://raw.githubusercontent.com/briatte/ggcorr/master/ggcorr.R') -->
<!-- # library(ggplot2) -->
<!-- #  -->
<!-- # ggcorr(xm) -->
<!-- # ggcorr(xm, nbreaks=3) -->
<!-- #  -->
<!-- # ggcorr(xm, geom = "blank", label = TRUE, hjust = 0.75) + -->
<!-- #   geom_point(size = 10, aes(color = coefficient > 0, alpha = abs(coefficient) > 0.5)) + -->
<!-- #   scale_alpha_manual(values = c("TRUE" = 0.25, "FALSE" = 0)) + -->
<!-- #   guides(color = FALSE, alpha = FALSE) -->

## Compare MSPE of Ridge and LASSO on air pollution data

Since our focus was on the properties of the resulting predictions, it may be
interesting to compare the estimated MSPE of the different models / predictors
we have considered so far: a full linear model, a model selected via stepwise + AIC, 
ridge regression and LASSO. As usual, we use 50 runs of 5-fold CV, and obtain
the following boxplots:

```{r bigcompare, fig.width=5, fig.height=5, warning=FALSE, message=FALSE, echo=FALSE}
library(MASS)
n <- nrow(xm)
k <- 5
ii <- (1:n) %% k + 1
set.seed(123)
N <- 50
mspe.la <- mspe.st <- mspe.ri <- mspe.f <- rep(0, N)
for(i in 1:N) {
  ii <- sample(ii)
  pr.la <- pr.f <- pr.ri <- pr.st <- rep(0, n)
  for(j in 1:k) {
    tmp.ri <- cv.glmnet(x=xm[ii != j, ], y=y[ii != j], lambda=lambdas, 
                        nfolds=5, alpha=0, family='gaussian') 
    tmp.la <- cv.glmnet(x=xm[ii != j, ], y=y[ii != j], lambda=lambdas, 
                        nfolds=5, alpha=1, family='gaussian')
    null <- lm(MORT ~ 1, data=airp[ii != j, ])
    full <- lm(MORT ~ ., data=airp[ii != j, ])
    tmp.st <- stepAIC(null, scope=list(lower=null, upper=full), trace=FALSE)
    pr.ri[ ii == j ] <- predict(tmp.ri, s='lambda.min', newx=xm[ii==j,])
    pr.la[ ii == j ] <- predict(tmp.la, s='lambda.min', newx=xm[ii==j,])
    pr.st[ ii == j ] <- predict(tmp.st, newdata=airp[ii==j,])
    pr.f[ ii == j ] <- predict(full, newdata=airp[ii==j,])
  }
  mspe.ri[i] <- mean( (airp$MORT - pr.ri)^2 )
  mspe.la[i] <- mean( (airp$MORT - pr.la)^2 )
  mspe.st[i] <- mean( (airp$MORT - pr.st)^2 )
  mspe.f[i] <- mean( (airp$MORT - pr.f)^2 )
}
boxplot(mspe.la, mspe.ri, mspe.st, mspe.f, names=c('LASSO','Ridge', 'Stepwise', 'Full'), col=c('steelblue', 'gray80', 'tomato', 'springgreen'), cex.axis=1, cex.lab=1, cex.main=2)
mtext(expression(hat(MSPE)), side=2, line=2.5)
```

We see that there is a marginal advantage of LASSO, but it is rather minor, and 
the three methods we have seen so far improve by similar margins 
on the predictions obtained by
using a full linear regression model. 

## Less desirable properties of LASSO

* Not "variable selection"-consistent
* Not oracle
* Tends to only pick one variable (randomly) from groups of correlated ones 

## Elastic net

Elastic Net estimators were introduced to find an 
informative compromise between LASSO and Ridge Regression. 

Note that `cv.glmnet` only considers fits with variying 
values of one of the penalty constants, while the other
one (`alpha`) is kept fixed. To compare different 
Elastic Net fits we run `cv.glmnet` with 4 values of
`alpha`: 0.05, 0.1, 0.5 and 0.75. 

```{r airp.en, fig.width=5, fig.height=5, warning=FALSE, message=FALSE}
# EN
set.seed(23)
air.en.75 <- cv.glmnet(x=xm, y=y, lambda=lambdas, nfolds=5, alpha=0.75, 
                 family='gaussian', intercept=TRUE)
set.seed(23)
air.en.05 <- cv.glmnet(x=xm, y=y, lambda=lambdas, nfolds=5, alpha=0.05, 
                 family='gaussian', intercept=TRUE)
set.seed(23)
air.en.1 <- cv.glmnet(x=xm, y=y, lambda=lambdas, nfolds=5, alpha=0.1, 
                 family='gaussian', intercept=TRUE)
set.seed(23)
air.en.5 <- cv.glmnet(x=xm, y=y, lambda=lambdas, nfolds=5, alpha=0.5, 
                 family='gaussian', intercept=TRUE)
plot(air.en.05)
plot(air.en.5)
plot(air.en.75)
```


### Run EN on airpollution data, compare fits

We now compare the estimates of the regression coefficients
obtained with the different methods discussed so far to 
alleviate potential problems caused by correlated covariates. 

```{r airp.en.compare}
a <- cbind(round(coef(air.l2, s='lambda.min'), 3),
round(coef(air.l1, s='lambda.min'), 3), 
round(coef(air.en.05, s='lambda.min'), 3),
round(coef(air.en.1, s='lambda.min'), 3),
round(coef(air.en.5, s='lambda.min'), 3), 
round(coef(air.en.75, s='lambda.min'), 3))
colnames(a) <- c('Ridge', 'LASSO', 'EN-05', 'EN-10', 'EN-50', 'EN-75')
a
```

The same comment made above regarding the need of a 
more stable choice of "optimal" fits (for each of these
methods) applies here. Again, here we limit ourselves to one
run of 5-fold CV purely based on simplifying the
presentation. 


### Compare MSPE's of Full, LASSO, Ridge, EN and stepwise


```{r bigcompare2, fig.width=5, fig.height=5, warning=FALSE, message=FALSE, echo=FALSE}
ii <- (1:n) %% k + 1
set.seed(123)
N <- 50
mspe.en <- rep(0, N)
for(i in 1:N) {
  ii <- sample(ii)
  pr.en <- rep(0, n)
  for(j in 1:k) {
    tmp.en <- cv.glmnet(x=xm[ii != j, ], y=y[ii != j], lambda=lambdas, 
                        nfolds=5, alpha=0.75, family='gaussian') 
    pr.en[ ii == j ] <- predict(tmp.en, s='lambda.min', newx=xm[ii==j,])
  }
  mspe.en[i] <- mean( (airp$MORT - pr.en)^2 )
}
boxplot(mspe.en, mspe.la, mspe.ri, mspe.st, mspe.f, 
        names=c('EN', 'LASSO','Ridge', 'Stepwise', 'Full'), 
        col=c('hotpink', 'steelblue', 'gray80', 'tomato', 'springgreen'), 
        cex.axis=1, cex.lab=1, cex.main=2)
mtext(expression(hat(MSPE)), side=2, line=2.5)
```

We see that in this example Elastic Net with `alpha = 0.75` (which is not far from 
the LASSO) provides slightly better estimated MSPEs.

<!-- # Non-parametric regression -->

<!-- ## Polynomial regression -->

<!-- ```{r nonparam, fig.width=5, fig.height=5, message=FALSE, warning=FALSE} -->
<!-- # help(lidar, package='SemiPar') -->

<!-- data(lidar, package='SemiPar') -->
<!-- plot(logratio~range, data=lidar, pch=19, col='gray', cex=1.5) -->

<!-- # Degree 4 polynomials -->
<!-- pm <- lm(logratio ~ poly(range, 4), data=lidar) -->
<!-- plot(logratio~range, data=lidar, pch=19, col='gray', cex=1.5) -->
<!-- lines(predict(pm)[order(range)] ~ sort(range), data=lidar, lwd=4, col='blue') -->

<!-- # Degree 10 polynomials -->
<!-- pm2 <- lm(logratio ~ poly(range, 10), data=lidar) -->
<!-- lines(predict(pm2)[order(range)]~sort(range), data=lidar, lwd=4, col='red') -->


<!-- # A more flexible basis: splines -->

<!-- # linear splines ``by hand'' -->
<!-- # select the knots at 5 quantiles -->
<!-- kn <- as.numeric( quantile(lidar$range, (1:5)/6) ) -->

<!-- # prepare the matrix of covariates / explanatory variables -->
<!-- x <- matrix(0, dim(lidar)[1], length(kn)+1) -->
<!-- for(j in 1:length(kn)) { -->
<!--   x[,j] <- pmax(lidar$range-kn[j], 0) -->
<!-- } -->
<!-- x[, length(kn)+1] <- lidar$range -->

<!-- # Fit the regression model -->
<!-- ppm <- lm(lidar$logratio ~ x) -->
<!-- plot(logratio~range, data=lidar, pch=19, col='gray', cex=1.5) -->
<!-- lines(predict(ppm)[order(range)]~sort(range), data=lidar, lwd=4, col='green') -->

<!-- # a better way to obtain the same fit -->
<!-- library(splines) -->
<!-- ppm2 <- lm(logratio ~ bs(range, degree=1, knots=kn), data=lidar) -->
<!-- lines(predict(ppm)[order(range)]~sort(range), data=lidar, lwd=2, col='blue') -->

<!-- # quadratic splines? -->
<!-- plot(logratio~range, data=lidar, pch=19, col='gray', cex=1.5) -->
<!-- ppmq <- lm(logratio ~ bs(range, degree=2, knots=kn), data=lidar) -->
<!-- lines(predict(ppmq)[order(range)]~sort(range), data=lidar, lwd=4, col='steelblue') -->

<!-- # cubic splines -->
<!-- plot(logratio~range, data=lidar, pch=19, col='gray', cex=1.5) -->
<!-- ppmc <- lm(logratio ~ bs(range, degree=3, knots=kn), data=lidar) -->
<!-- lines(predict(ppmc)[order(range)]~sort(range), data=lidar, lwd=4, col='tomato3') -->
<!-- ``` -->
