---
title: "STAT406 - Lecture 12 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

The lecture slides are [here](STAT406-17-lecture-12-preliminary.pdf).


## Bagging

One strategy to obtain more stable predictors is called
**Bootstrap AGGregatING** (bagging). It can be applied to
many predictors (not only trees), and it generally results
in larger improvements in prediction quality when it is used with predictors
that are flexible (low bias), but highly variable.

The justification and motivation were discussed in class. Intuitively
we are averaging the predictions obtained from an estimate of the 
"average prediction" we would have computed had we had access to 
several (many?) independent training sets (samples). 

There are several (many?) `R` packages implementing
bagging for different predictors, with varying degrees of 
flexibility (the implementations) and user-friendliness. 
However, for pedagogical and illustrative purposes, in these notes I will
*bagg* by hand.

### Bagging by hand

Again, to simplify the discussion and presentation, in order to evaluate 
prediction quality I will split the 
data (`Boston`) into a training and a test set. We do this now:
```{r bag1, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
library(rpart)
data(Boston, package='MASS')
set.seed(123456)
n <- nrow(Boston)
ii <- sample(n, floor(n/4))
dat.te <- Boston[ ii, ]
dat.tr <- Boston[ -ii, ]
```
I will now train $N = 5$ trees and average their predictions. 
Note that, in order to illustrate the process more
clearly, I will compute and store the $N \times n_e$
predictions, where $n_e$ denotes the number of observations in 
the test set. This is not the best (most efficient) way of implementing *bagging*,
but the main purpose here is to understand **what** we are doing. Also note that
an alternative (better in terms of reusability of the
ensamble, but maybe still not the most efficient option) would be
to store the $N$ trees directly. This would also allow for
more elegant and easy to read code. Once again, this approach 
will be sacrificed in the altar of clarity of presentation and 
pedagogy (but do try it yourself!)

First create an array where we will store all the predictions:
```{r bag2, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
N <- 5
myps <- array(NA, dim=c(nrow(dat.te), N))
con <- rpart.control(minsplit=3, cp=1e-3, xval=1)
```
The last object (`con`) contains my options to train large
(potentially overfitting) trees. 
```{r bag3, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
And compare with predictions from the pruned tree, and the
ones from other predictors discussed in the previous note:
```{r bag4, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
myc <- rpart.control(minsplit=3, cp=1e-8, xval=10)
set.seed(123)
bos.to <- rpart(medv ~ ., data=dat.tr, method='anova',
                control=myc)
b <- bos.to$cptable[which.min(bos.to$cptable[,"xerror"]),"CP"]
bos.t3 <- prune(bos.to, cp=b)
pr.t3 <- predict(bos.t3, newdata=dat.te, type='vector')
with(dat.te, mean((medv - pr.t3)^2) )
```

What if we *bagg* $N = 10$ trees? 
```{r bag10, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 10
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
or $N = 100$ trees? 
```{r bag100, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 100
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
or $N = 1000$ trees? 
```{r bag1000, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 1000
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```

Should we consider higher values of $N$? How about other
training / test splits? Should we use CV instead? 

Another split:
```{r anothersplit, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
set.seed(123)
n <- nrow(Boston)
ii <- sample(n, floor(n/4))
dat.te <- Boston[ ii, ]
dat.tr <- Boston[ -ii, ]
for(N in c(5, 10, 100, 1000)) {
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
print(c(N, with(dat.te, mean( (medv - pr.bagg)^2 ) )))
}
```

Similar conclusion: increasing $N$ helps, but the improvement 
becomes smaller, while the computational cost keeps increasing. 

### Bagging a regression spline

Bagging does not provide much of an advantage when applied to linear
predictors (can you explain why?) Nevertheless, let us try it on the `lidar` data, 
which, as we did before, we randomly split into a training and test set:
```{r bagsplines, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
data(lidar, package='SemiPar')
set.seed(123456)
n <- nrow(lidar)
ii <- sample(n, floor(n/5))
lid.te <- lidar[ ii, ]
lid.tr <- lidar[ -ii, ]
```
Now fit a cubic spline, and estimate the MSPE using the test set:
```{r bagsplines2, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
library(splines)
a <- lm(logratio ~ bs(x=range, df=10, degree=3), data=lid.tr) 
oo <- order(lid.tr$range)
pr.of <- predict(a, newdata=lid.te)
mean( (lid.te$logratio - pr.of)^2 )
```
We build an ensemble of 10 fits and estimate the corresponding
MSPE using the test set:
```{r bagsplines3, fig.width=5, fig.height=5, message=FALSE, warning=FALSE}
N <- 10 # 5 500 1500
myps <- matrix(NA, nrow(lid.te), N)
set.seed(123456)
n.tr <- nrow(lid.tr)
for(i in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  a.b <- lm(logratio ~ bs(x=range, df=10, degree=3), data=lid.tr[ii,]) 
  myps[,i] <- predict(a.b, newdata=lid.te)
}
pr.ba <- rowMeans(myps)# , na.rm=TRUE)
mean( (lid.te$logratio - pr.ba)^2 )
```


## Linear Discriminant Analysis

As we discussed in class, what is commonly referred to as *classification* 
can be thought of as prediction, when the responses are classes and we
use a particular loss function (the *0-1 loss* we discussed in class). Most
(if not all) classification methods we will cover in this course are 
simply different approaches to estimate the conditional probability of
each class, conditional on the value of the explanatory variables. In 
symbols: $P( \mathbf{G} = \mathbf{g} | \mathbf{X} = \mathbf{x}_0)$.
The obvious parallel in what we have done before this Term,
is that many (all?) regression methods are 
different ways of estimating the conditional mean of the response 
variable, conditional on the value of the features (explanatory
variables). 

Probably the "second easiest to explain" approach (what is "*the* easiest 
to explain" then?) to estimate the above probability is to model the distribution of 
the explanatory variables *within* each class. This will then uniquely determine the 
probabilities we need to estimate, as discussed above and in class. 
In particular, one the simplest models we can use is a normal (Gaussian) one. 
As we saw in class, if we assume that the distribution of the features
for each class is Gaussian with a common covariance matrix across clases, then
it easy to show (*and I strongly suggest that you do it*) that the optimal
classifier (using the 0-1 loss function m.entioned above) is a linear
function of the explanatory variables. The coefficients of this linear
function depend on the parameters of the assumed Gaussian distributions. The
latter can be estimated using the training set and plugging them in provides
a natural estimator of the optimal classifier. 

The function `lda` in 
the `MASS` library implements this simple classifier. We illustrate it
here on the rather simple and well-known vaso constriction data, available in the 
`robustbase` package. More details, as usual, can be found on its
help page. The response variable takes two values (represented below
as **blue** and **red**), and there are only two
explanatory variables (which allows us to visualize our methods and results). 

```{r lda1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
data(vaso, package='robustbase')
plot(Volume ~ Rate, data=vaso, pch=19, cex=1.5, 
       col=c('red', 'blue')[Y+1])
```

To train the LDA classifier we use the function `lda` as follows (note the
*model-like* syntax to indicate the response and explanatory variables):

```{r lda1.1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(MASS)
a.lda <- lda(Y ~ Volume + Rate, data=vaso)
```

Now, given any value of the explanatory variables `(Volume, Rate)` we
can use the method `predict` on the object returned by `lda()` to 
estimate the conditional probabilities of **blue** and **red**. 
Next we visualize which regions of the feature space will be predicted to 
contain **blue** points (and then obviously which areas will be 
predicted to correspond to **red** responses). To do this we 
build a relatively fine grid of posible `(Volume, Rate)` values
and plot the corresponding predictions:

```{r lda1.2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
# build a grid of points of (Volume, Rate),
# and obtain predictions for each of them
# 40,000 in total
xvol <- seq(0, 4, length=200)
xrat <- seq(0, 4, length=200)
xx <- expand.grid(xvol, xrat)
names(xx) <- c('Volume', 'Rate')

pr.lda <- predict(a.lda, newdata=xx)$posterior[,2]
image(xrat, xvol, matrix(pr.lda, 200, 200), col=terrain.colors(100),
      ylab='Volume', xlab='Rate', main='LDA')
points(Volume ~ Rate, data=vaso, pch=19, cex=1.5,
       col=c('red', 'blue')[Y+1])
```

This model-based approach to classification can be relaxed slightly if
we assume instead that the features have a Gaussian distribution within each
class, but that the covariance matrix may be different across 
classes. The corresponding optimal classifier is now a quadratic
function of the predictors (**prove it!**). The function `qda` 
in the `MASS` library implements this classifier.  

This approach can be used with any number of classes. Any limitations? 

## Logistic regression (Review)

If we model the distribution of the  features within each class using a 
multivariate Gaussian distribution, then it is easy to see that the 
boundaries between classes are linear (functions of the features) (*check!*)
Furthermore, the log of the odds ratio between classes is a linear
function. It is interesting to note that one can start with this last
assumption (instead of the full Gaussian model) and arrive at a
fully parametric model for the conditional distibution of the classes
given the features (see the class slides). The parameters can be
estimated using maximum likelihood. For two classes this is the
logistic regression model, which you may have seen in previous
courses. 

We illustrate this on the `vaso` data as before. Since this is
a 2-class problem, we just need to fit a logistic regression model.
The function `glm` in `R` does it for us, we specify that we 
want to fit such a model using the argument `family=binomial`. 
Once we obtain parameter estimators (in the `glm` object `a` below), 
we use the `predict` method to obtain predicted conditional 
probabilities on the same grid we used before:
```{r logistic1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
a <- glm(Y ~ ., data=vaso, family=binomial)
pr <- predict(a, newdata=xx, type='response')
```

We now plot the data and the *surface* of predicted probabilities for 
blue points (higher probabilites are displayed with lighter colors). 

```{r logistic2, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
image(xrat, xvol, matrix(pr, 200, 200), col=terrain.colors(100),
      ylab='Volume', xlab='Rate', main='Logistic')
points(Volume ~ Rate, data=vaso, pch=19, cex=1.5,
       col=c('red', 'blue')[Y+1])
```

