---
title: "STAT406 - Lecture 12 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

The lecture slides are [here](STAT406-17-lecture-12-preliminary.pdf).


## Bagging

One strategy to obtain more stable predictors is called
**Bootstrap AGGregatING** (bagging). It can be applied to
many predictors (not only trees), and it generally results
in larger improvements in prediction quality when it is used with predictors
that are flexible (low bias), but highly variable.

The justification and motivation were discussed in class. Intuitively
we are averaging the predictions obtained from an estimate of the 
"average prediction" we would have computed had we had access to 
several (many?) independent training sets (samples). 

There are several (many?) `R` packages implementing
bagging for different predictors, with varying degrees of 
flexibility (the implementations) and user-friendliness. 
However, for pedagogical and illustrative purposes, in these notes I will
*bagg* by hand.

### Bagging by hand

Again, to simplify the discussion and presentation, in order to evaluate 
prediction quality I will split the 
data (`Boston`) into a training and a test set. We do this now:
```{r bag1, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
library(rpart)
data(Boston, package='MASS')
set.seed(123456)
n <- nrow(Boston)
ii <- sample(n, floor(n/4))
dat.te <- Boston[ ii, ]
dat.tr <- Boston[ -ii, ]
```
I will now train $N = 5$ trees and average their predictions. 
Note that, in order to illustrate the process more
clearly, I will compute and store the $N \times n_e$
predictions, where $n_e$ denotes the number of observations in 
the test set. This is not the best (most efficient) way of implementing *bagging*,
but the main purpose here is to understand **what** we are doing. Also note that
an alternative (better in terms of reusability of the
ensamble, but maybe still not the most efficient option) would be
to store the $N$ trees directly. This would also allow for
more elegant and easy to read code. Once again, this approach 
will be sacrificed in the altar of clarity of presentation and 
pedagogy (but do try it yourself!)

First create an array where we will store all the predictions:
```{r bag2, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
N <- 5
myps <- array(NA, dim=c(nrow(dat.te), N))
con <- rpart.control(minsplit=3, cp=1e-3, xval=1)
```
The last object (`con`) contains my options to train large
(potentially overfitting) trees. 
```{r bag3, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
And compare with predictions from the pruned tree, and the
ones from other predictors discussed in the previous note:
```{r bag4, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
myc <- rpart.control(minsplit=3, cp=1e-8, xval=10)
set.seed(123)
bos.to <- rpart(medv ~ ., data=dat.tr, method='anova',
                control=myc)
b <- bos.to$cptable[which.min(bos.to$cptable[,"xerror"]),"CP"]
bos.t3 <- prune(bos.to, cp=b)
pr.t3 <- predict(bos.t3, newdata=dat.te, type='vector')
with(dat.te, mean((medv - pr.t3)^2) )
```

What if we *bagg* $N = 10$ trees? 
```{r bag10, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 10
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
or $N = 100$ trees? 
```{r bag100, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 100
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```
or $N = 1000$ trees? 
```{r bag1000, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
N <- 1000
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
with(dat.te, mean( (medv - pr.bagg)^2 ) )
```

Should we consider higher values of $N$? How about other
training / test splits? Should we use CV instead? 

Another split:
```{r anothersplit, fig.width=5, fig.height=5, message=FALSE, warning=TRUE, echo=FALSE}
set.seed(123)
n <- nrow(Boston)
ii <- sample(n, floor(n/4))
dat.te <- Boston[ ii, ]
dat.tr <- Boston[ -ii, ]
for(N in c(5, 10, 100, 1000)) {
myps <- array(NA, dim=c(nrow(dat.te), N))
n.tr <- nrow(dat.tr)
set.seed(123456)
for(j in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  tmp <- rpart(medv ~ ., data=dat.tr[ii, ], method='anova', control=con)
  myps[,j] <- predict(tmp, newdata=dat.te, type='vector')
}
pr.bagg <- rowMeans(myps)
print(c(N, with(dat.te, mean( (medv - pr.bagg)^2 ) )))
}
```

Similar conclusion: increasing $N$ helps, but the improvement 
becomes smaller, while the computational cost keeps increasing. 

### Bagging a regression spline

Bagging does not provide much of an advantage when applied to linear
predictors (can you explain why?) Nevertheless, let us try it on the `lidar` data, 
which, as we did before, we randomly split into a training and test set:
```{r bagsplines, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
data(lidar, package='SemiPar')
set.seed(123456)
n <- nrow(lidar)
ii <- sample(n, floor(n/5))
lid.te <- lidar[ ii, ]
lid.tr <- lidar[ -ii, ]
```
Now fit a cubic spline, and estimate the MSPE using the test set:
```{r bagsplines2, fig.width=5, fig.height=5, message=FALSE, warning=TRUE}
library(splines)
a <- lm(logratio ~ bs(x=range, df=10, degree=3), data=lid.tr) 
oo <- order(lid.tr$range)
pr.of <- predict(a, newdata=lid.te)
mean( (lid.te$logratio - pr.of)^2 )
```
We build an ensemble of 10 fits and estimate the corresponding
MSPE using the test set:
```{r bagsplines3, fig.width=5, fig.height=5, message=FALSE, warning=FALSE}
N <- 10 # 5 500 1500
myps <- matrix(NA, nrow(lid.te), N)
set.seed(123456)
n.tr <- nrow(lid.tr)
for(i in 1:N) {
  ii <- sample(n.tr, replace=TRUE)
  a.b <- lm(logratio ~ bs(x=range, df=10, degree=3), data=lid.tr[ii,]) 
  myps[,i] <- predict(a.b, newdata=lid.te)
}
pr.ba <- rowMeans(myps)# , na.rm=TRUE)
mean( (lid.te$logratio - pr.ba)^2 )
```


## Linear Discriminant Analysis

```{r lda1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
library(MASS)
data(vaso, package='robustbase')
a.lda <- lda(Y ~ Volume + Rate, data=vaso)

# build a grid of points of (Volume, Rate),
# and obtain predictions for each of them
# 40,000 in total
xvol <- seq(0, 4, length=200)
xrat <- seq(0, 4, length=200)
xx <- expand.grid(xvol, xrat)
names(xx) <- c('Volume', 'Rate')

pr.lda <- predict(a.lda, newdata=xx)$posterior[,2]
image(xrat, xvol, matrix(pr.lda, 200, 200), col=terrain.colors(100),
      ylab='Volume', xlab='Rate', main='LDA')
points(Volume ~ Rate, data=vaso, pch=19, cex=1.5,
       col=c('red', 'blue')[Y+1])
```


## Logistic regression (Review)

```{r logistic1, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
plot(Volume ~ Rate, data=vaso, pch=19, cex=1.5, col=c('red', 'blue')[Y+1],
     xlim=c(0, 4), ylim=c(0,4))

a <- glm(Y ~ ., data=vaso, family=binomial)

pr <- predict(a, newdata=xx, type='response')

# display them
image(xrat, xvol, matrix(pr, 200, 200), col=terrain.colors(100),
      ylab='Volume', xlab='Rate', main='Logistic')
points(Volume ~ Rate, data=vaso, pch=19, cex=1.5,
       col=c('red', 'blue')[Y+1])
# Y = 1 corresponds to blue points
# higher probabilities are displayed with lighter colors
```

