---
title: "DSCI563 - Lecture 8 notes"
author: "Matias Salibian"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lecture slides 

The lecture slides are [here](DSCI563-16-17-lecture-8.pdf) and
[here](Assoc-Rules.pdf) (the second set courtesy of Mark Schmidt).

## Multidimensional scaling

#### European languages example

We will illustrate it first on the languages example we
saw earlier in class. We have a specific dissimilarity
measure between 11 european languages (the number of
different first letters in their numerals). Recall
what we found when use used a hierarchical clustering 
algorithm:

```{r lang1, fig.width=5, fig.height=5, echo=FALSE,  message=FALSE, warning=FALSE}
a.la <- read.table('languages.dat', header=FALSE)
a.la <- a.la + t(a.la)
la.hc <- hclust(as.dist(a.la), method='ward.D')
la.nms <- c(
'E',
'N',
'Da',
'Du',
'G',
'Fr',
'S',
'I',
'P',
'H',
'Fi'
)
plot(la.hc, labels=la.nms)
```

We will now use Classical (Metric) Multidimensional Scaling
as implemented in the `R` function `cmdscale` to find a 
two dimensional representation of these countries that
best reflects their pairwise distances

```{r lang2, fig.width=5, fig.height=5, echo=TRUE,  message=FALSE, warning=FALSE}
a.la <- read.table('languages.dat', header=FALSE)
a.la <- a.la + t(a.la)
la.cmds <- cmdscale(a.la, k=2)
plot(la.cmds, type='n')
text(la.cmds, labels=la.nms)
```

Note that the 2-dimensional representation above does not match 
very well
what the hierarchical clustering dendongram
suggests (e.g. **N** appears closer to **G** than to **Da**, **E** appears
closer to **Da** than **N**, etc.) 
We now try using the more flexible isotonic (non-metric method)
as implemented in the function `isoMDS` of package `MASS`:

```{r lang3, fig.width=5, fig.height=5, echo=TRUE,  message=FALSE, warning=FALSE}
la.iso <- MASS::isoMDS(as.matrix(a.la), k=2, maxit=500, tol=1e-6)
plot(la.iso$points, type='n')
text(la.iso$points, labels=la.nms)
```

This representation matches more closely what we found with the
hierarchical clustering analysis, and seems intuitively better. 

We can now try ISOMAP, which is implemented in the function 
`isomap` in the `vegan` package for `R`:
```{r lang4, fig.width=5, fig.height=5, echo=TRUE, message=FALSE, warning=FALSE}
library(vegan)
a <- isomap(a.la, ndim=2, k=3)
plot(a, type='points', net=TRUE) 
text(a$points, labels=la.nms, pos=3, offset=.3)
```

Note that the ISOMAP result is very close to the `isoMDS` solution. 

#### UN votes example

We now apply these dimension-reduction methods to the 
UN votes data. First use the classical / metric MDS approach 
to find a lower-dimensional representation
of the countries in such a way that the pairwise
distances are roughly preserved (to the extent that it is possible):
```{R initial, tidy=TRUE}
X <- read.table(file='../Lecture2-Hierarchical/unvotes.csv', sep=',', row.names=1, header=TRUE)
X2 <- X[complete.cases(X),]
dd <- dist(t(X2)) 
tmp <- cmdscale(dd, k =2) 
```

The result is:

```{r scatter, fig.width=8, fig.height=8, echo=FALSE,  message=FALSE, warning=FALSE}
plot(tmp, pch=19, col='gray50', cex=2, type='n', xlab='', ylab='')
text(tmp[,1], tmp[,2], labels=rownames(tmp), cex=.8)
```

Using ISOMAP we get:

```{r unvotes, fig.width=8, fig.height=8, echo=FALSE,  message=FALSE, warning=FALSE}
a <- isomap(dd, ndim=2, k=10)
plot(a, type='text', net=TRUE) 
```

#### Digits example (using Euclidean distances)

Recall the 
digitized hand-written "3"s example:
```{r three1}
a <- read.table('train.3.txt', header=FALSE, sep=',')
```
There are 658 images, each of them represented
on a grid of 16x16 pixels (the data are pixel intensities
scaled to be between -1 and 1).

```{r oneimage, fig.width=5, fig.height=5, echo=FALSE}
myImagePlot <- function(x){
  min <- min(x)
  max <- max(x)
  ColorRamp <- grey(seq(1,0,length=256))
  ColorLevels <- seq(min, max, length=length(ColorRamp))
  reverse <- nrow(x):1
  x <- x[reverse,]
  image(1:ncol(x), 1:nrow(x), t(x), col=ColorRamp, xlab="",
        ylab="", axes=FALSE, zlim=c(min,max))
}
n <- dim(a)[1]
set.seed(31)
myImagePlot(t(matrix(unlist(a[sample(n,1),]), 16, 16)))
```

We now use ISOMAP to find a 2-dimensional representation,
and use it to explore these images
```{r three2, fig.width=5, fig.height=5, echo=TRUE,  message=FALSE, warning=FALSE}
dd <- dist(a)
# dd <- 1-abs(cor(t(a)))
a.iso <- isomap(dd, ndim=2, k=5)
plot(a.iso, type='points', net=TRUE) 
```

We now "walk" along the manifold in different
directions and can look at the corresponding images. 
The points were identified interactively using 
the function `identify` in `R`:

```{r three3, fig.width=5, fig.height=5, echo=FALSE,  message=FALSE, warning=FALSE}
plot(a.iso, type='points', net=TRUE) 
sp <- c(116, 153, 172, 287, 371, 489)
points(a.iso$points[sp, ], pch=19, col='red')
```

We now overlay the corresponding images

```{r three4, fig.width=8, fig.height=8, echo=FALSE, message=FALSE, warning=FALSE}
plot(a.iso, type='points', net=TRUE) 
library(fields)
for(j in sp) {
  im <- matrix(unlist(a[j,]), 16, 16)
  add.image(a.iso$points[j,1], a.iso$points[j,2], im, image.width=.1, col=grey(seq(1,0,length=256)))
}
```

We now do the same along a different line

```{r three5, fig.width=8, fig.height=8, echo=FALSE, message=FALSE, warning=FALSE}
plot(a.iso, type='points', net=TRUE) 
sp <- c(220, 243, 324, 353, 454, 517)
for(j in sp) {
  im <- matrix(unlist(a[j,]), 16, 16)
  add.image(a.iso$points[j,1], a.iso$points[j,2], im, image.width=.1, col=grey(seq(1,0,length=256)))
}
```

#### Digits example (using correlation-based dissimilarities)

The above exploratory experiment wasn't as successful as the
one we did based on principal components. Below we 
repeat it using a different dissimilarity measure
between images, in the hope of finding a more interesting
pattern in the data. 

Another dissimilarity measure that can be used is based
on correlations. Specifically we define the dissimilarity d(**a**, **b**) between
points **a** and **b**
as d(**a**, **b**) = 1 - abs(cor(a, b)), so that 
d(**a**, **b**) = 0 when **a** and **b** are perfectly linearly 
correlated, an 
d(**a**, **b**) is maximal when there is no linear correlation
between the entries (elements) of **a** and **b**.

We will re-do the ISOMAP analysis above on the hand-written
digits, but using this dissimilarity measure. Since 
images are stored row-wise in the matrix `a` in the
`R` code above, to compute
the "correlations" between them I will use the function 
`cor` but applied to the transposed data matrix:

```{r three6, fig.width=5, fig.height=5, echo=TRUE,  message=FALSE, warning=FALSE}
dd <- 1-abs(cor(t(a)))
a.iso <- isomap(dd, ndim=2, k=5)
plot(a.iso, type='points', net=TRUE) 
```

We again use the function `identify` to select a few
images to reveal 
along different paths: 

```{r three7, fig.width=8, fig.height=8, echo=TRUE,  message=FALSE, warning=FALSE}
plot(a.iso, type='points', net=TRUE) 
# sp <- identify(a.iso$points)
sp <- c(9,  46, 101, 330, 343, 381, 396, 626)
for(j in sp) {
  im <- matrix(unlist(a[j,]), 16, 16)
  add.image(a.iso$points[j,1], a.iso$points[j,2], im, image.width=.1, col=grey(seq(1,0,length=256)))
}
```

```{r three8, fig.width=8, fig.height=8, echo=TRUE,  message=FALSE, warning=FALSE}
plot(a.iso, type='points', net=TRUE) 
# sp <- identify(a.iso$points)
sp <- c(51,  99, 104, 155, 336, 400, 438, 521)
for(j in sp) {
  im <- matrix(unlist(a[j,]), 16, 16)
  add.image(a.iso$points[j,1], a.iso$points[j,2], im, image.width=.1, col=grey(seq(1,0,length=256)))
}
```

Unfortunately no easily interpretable pattern is apparent either. 



