---
title: "STAT406 - Lecture 24 notes"
author: "Matias Salibian-Barrera"
date: "`r format(Sys.Date())`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=6, fig.height=6,  
                      warning=FALSE, message=FALSE, tidy=TRUE) #, eval = FALSE)
```

## LICENSE
These notes are released under the 
"Creative Commons Attribution-ShareAlike 4.0 International" license. 
See the **human-readable version** [here](https://creativecommons.org/licenses/by-sa/4.0/)
and the **real thing** [here](https://creativecommons.org/licenses/by-sa/4.0/legalcode). 

## Lecture slides

The lecture slides are [here](STAT406-17-lecture-24-preliminary.pdf).


## Principal Components Analysis

Although principal components can be easily computed with the spectral
decomposition of the covariance matrix of the data (using the function 
`svd` in `R`, for example), there are a few dedicated implementations
in `R`, among them `prcomp` and `princomp`). The main difference between
these two is which internal function is used to compute eigenvalues and
eigenvectors: `prcomp` uses `svd` and `princomp` uses the less preferred
function `eigen`. Both `princomp` and `prcomp` return the matrix of 
loadings (eigenvectors), the scores (projections of the data on the 
basis of eigenvectors), and other auxiliary objects. They also include
plot and summary methods.

Instead of reviewing those (which can be easily done individually), in
these notes I will reproduce two of the examples used in class (the simple
2-dimensional one used to motivate the topic, and the more interesting
256-dimensional one using the digits data). Finally, I will also 
show that principal components can be computed using an iterative
algorithm (alternate regression), which may be faster than factorizing
the covariance matrix, particularly when one is only interested in a
few principal components and the dimension of the data is large. 


#### Example 1

Read the data for the simple illustration of PC's as best lower 
dimensional approximations
```{r pca1}
x <- read.table('t8-5.dat', header=TRUE)
```
We will only use two coordinates in order to be able to use plots
```{r pca2}
xx <- x[,c(2,5)]
colnames(xx) <- c('Prof degree', 'Median home value')
```
We now standardize the data
```{r pca3}
xx <- scale(xx, center=colMeans(xx), scale=TRUE)
```
The following are auxiliary functions to compute Euclidean
norms and squared Euclidean norms (faster than `R`'s `norm`):
```{r norms}
norm <- function(a) sqrt(norm2(a))
norm2 <- function(a) sum(a^2)
```

A scatter plot of the data

```{r data}
plot(xx, pch=19, col='black', cex=2, xlim=c(-1.5, 4.5), ylim=c(-1.5, 4.5))
abline(h=0, lwd=2, col='grey', lty=2)
abline(v=0, lwd=2, col='grey', lty=2)
```

We now compute projections along the direction of the vector
(1, 0.05)':

```{r proj1}
a <- c(1, 0.05)
a <- a / norm(a)
# Find the projections (coordinates of the
# observations on this basis of size 1)
prs <- (xx %*% a)
# Find the orthogonal projections of each
# observation on this subspace of dimension 1
pr <- prs %*% a 
# Plot the data
plot(xx, pch=19, col='black', cex=2, xlim=c(-1.5, 4.5), ylim=c(-1.5, 4.5))
abline(h=0, lwd=2, col='grey', lty=2)
abline(v=0, lwd=2, col='grey', lty=2)

# Show the subspace on which we are projecting
abline(0, a[2]/a[1], lwd=2, col='red')

# Add the projections of the data on this subspace
points(pr[,1], pr[,2], pch=19, cex=1.5, col='red')

# Highlight a few of them
ind <- c(26, 25, 48, 36)
pr2 <- pr[ind, ]
for(j in 1:length(ind))
lines(c(xx[ind[j],1], pr2[j,1]), c(xx[ind[j],2], pr2[j,2]), 
col='blue', lwd=3.5, lty=2)
```

We can repeat for a different direction, e.g. (-1, 3)': 

```{r proj2, echo=FALSE}
a <- c(-1, 3)
a <- a / norm(a)
# Find the projections (coordinates of the
# observations on this basis of size 1)
prs <- (xx %*% a)
# Find the orthogonal projections of each
# observation on this subspace of dimension 1
pr <- prs %*% a 
# Plot the data
plot(xx, pch=19, col='black', cex=2, xlim=c(-1.5, 4.5), ylim=c(-1.5, 4.5))
abline(h=0, lwd=2, col='grey', lty=2)
abline(v=0, lwd=2, col='grey', lty=2)

# Show the subspace on which we are projecting
abline(0, a[2]/a[1], lwd=2, col='red')

# Add the projections of the data on this subspace
points(pr[,1], pr[,2], pch=19, cex=1.5, col='red')

# Highlight a few of them
ind <- c(26, 25, 48, 36)
pr2 <- pr[ind, ]
for(j in 1:length(ind))
lines(c(xx[ind[j],1], pr2[j,1]), c(xx[ind[j],2], pr2[j,2]), 
col='blue', lwd=3.5, lty=2)
```


#### Digits example

Read the data
```{r digits1}
a <- read.table('train.3.txt', header=FALSE, sep=',')
# Verify that the data has been read correctly
dim(a)
```
Define auxiliary functions for the squared Euclidean norm
and the squared Euclidean distance
```{r aux2}
norm2 <- function(u) sum(u^2)
dist <- function(a, b) norm2(a-b)
```
```{r plotfunction, echo=FALSE}
# ----- Define a function for plotting a matrix ----- #
# modified from: http://www.phaget4.org/R/image_matrix.html
# ----- Define a function for plotting a matrix ----- #
myImagePlot <- function(x, ...){
     min <- min(x)
     max <- max(x)
 ColorRamp <- grey(seq(1,0,length=256))
 ColorLevels <- seq(min, max, length=length(ColorRamp))
 # Reverse Y axis
 reverse <- nrow(x) : 1
 x <- x[reverse,]
image(1:ncol(x), 1:nrow(x), t(x), col=ColorRamp, xlab="",
 ylab="", axes=FALSE, zlim=c(min,max))
}
```

Plot 9 random chosen images (hand-written 3's).

```{r digitsplot}
set.seed(31)
sa <- sample(dim(a)[1], 9)
par(mai = c(1, 1, 1, 1)/5, xaxs = "i", yaxs = "i")
par(mfrow=c(3,3))
for(j in 1:9) {
myImagePlot(t(matrix(unlist(a[sa[j],]), 16, 16)))
}
par(mfrow=c(1,1))
```

<!-- # plot the average 3 -->
<!-- myImagePlot(t(matrix(colMeans(a), 16, 16))) -->


<!-- # Plot the first 3 on the data set -->
<!-- myImagePlot(t(matrix(unlist(a[1,]), 16, 16))) -->

Centre the observations, to compute the eigenvectors more efficiently
```{r digits3}
ac <- scale(a, center=TRUE, scale=FALSE)
si.svd <- svd(ac)
```
Compute the proportion of the "total variance" that is accounted by 
the first 20 ppal components:
```{r digits4}
si.svd$d[1:20]^2 / sum(si.svd$d^2)
# cumulative proportions:
cumsum(si.svd$d^2)[1:20] / sum(si.svd$d^2)
```
Compute the projections of the centered data on each of the 1st, 2nd and 3rd eigenvectors
```{r digits.p1}
v1 <- as.vector( ac %*% si.svd$v[,1] )
v2 <- as.vector( ac %*% si.svd$v[,2] )
v3 <- as.vector( ac %*% si.svd$v[,3] )
```
Find 5 quantiles of each of these coordinates
```{r digits.p2}
qv1 <- quantile(v1, c(.05, .25, .5, .75, .95))
qv2 <- quantile(v2, c(.05, .25, .5, .75, .95))
```
Plot the 2-dimensional projections plus the 5x5 grid on top

```{r digits.p3}
# Create a grid
qv <- expand.grid(qv1, qv2)
plot(v1, v2, pch=19, cex=1, col='grey')
points(qv[,1], qv[,2], pch=19, cex=1.5, col='red')
```

Now find the (projected) points closest to each of the 5x5=25 points in the grid
```{r digits.p4}
vs <- cbind(v1, v2)
cvs <- array(0, dim=dim(qv))
for(j in 1:dim(qv)[1]) {
cvs[j,] <- vs[ which.min( apply(vs, 1, dist, b=qv[j,]) ), ]
}
```
Identify these points in blue:

```{r digits.p5}
plot(v1, v2, pch=19, cex=1, col='grey')
points(qv[,1], qv[,2], pch=19, cex=1.5, col='red')
for(j in 1:dim(qv)[1]) {
points(cvs[j,1], cvs[j,2], pch=19, col='blue')
}
```

Find the approximations for these closest (blue) points (these will be points in a
2-dimensional subspace within R^256:
```{r digits.p6}
app <- t( si.svd$v[,1:2] %*% t( cvs ) )
```
Now find the images closest to these points
```{r digits.p7}
repre <- matrix(0, dim(qv)[1], dim(app)[2])
for(j in 1:dim(qv)[1]) {
repre[j,] <- ac[ which.min( apply(ac, 1, dist, b=app[j,]) ), ]
}
```
These are the actual images that are closest to the points in "app"
Now, add the column means and display these 25 images according to the points
they represent in the red grid:

```{r digits.p8}
repre <- scale(repre, center = -colMeans(a), scale=FALSE)
par(mai = c(1, 1, 1, 1)/5, xaxs = "i", yaxs = "i")
par(mfrow=c(5,5))
for(j in 1:dim(repre)[1]) {
myImagePlot(t(matrix(unlist(repre[j,]), 16, 16)))
}
par(mfrow=c(1,1))
```

Repeat with 3 ppal components now!:

```{r digits.3pca}
qv1 <- quantile(v1, c(.05, .25, .5, .75, .95))
qv2 <- quantile(v2, c(.05, .25, .5, .75, .95))
qv3 <- quantile(v3, c(.05, .5, .95))
qv <- expand.grid(qv1, qv2, qv3)

vs <- cbind(v1, v2, v3)
# closest v's
cvs <- array(0, dim=dim(qv))
for(j in 1:dim(qv)[1]) {
cvs[j,] <- vs[ which.min( apply(vs, 1, dist, b=qv[j,]) ), ]
}

app <- t( si.svd$v[,1:3] %*% t( cvs ) )


repre <- matrix(0, dim(qv)[1], dim(app)[2])

for(j in 1:dim(qv)[1]) {
repre[j,] <- ac[ which.min( apply(ac, 1, dist, b=app[j,]) ), ]
}

repre <- scale(repre, center = -colMeans(a), scale=FALSE)
par(mai = c(1, 1, 1, 1)/5, xaxs = "i", yaxs = "i")
par(mfrow=c(5,5))
for(j in 1:dim(repre)[1]) {
myImagePlot(t(matrix(unlist(repre[j,]), 16, 16)))
}
```
